Assessing variations in the chemistry of sub-fossil and modern *Pinus* pollen

### 2021-07-08 manuscript.Rmd
moving refs and removing whitespace

---
title: 'Assessing variations in the chemistry of fossil- and modern-*Pinus* pollen'
date: '2020/02/04'
output:
  bookdown::pdf_document2:
    toc: false
    extra_dependencies: ["flafter"]
  bookdown::word_document2:
bibliography: references.bib
csl: biology-letters.csl
---

@@

---
author: `r Authors`
---
`r Affiliations`

2500 words not including in tables title page references
2 figs and 2 tables
figures separate legends in text
separate cover letter

<!-- 
A  200
I  500
MM 400
R  500
D  800
-->

@@

The study of plant microfossils (pollen and spores) is an integral part of palaeoecology and palaeobotany.
Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g. the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001).

@@

Successful application of pollen-chemistry techniques within palaeoecological research requires a thorough understanding of the factors influencing chemical variations in the modern environment and in the accurate estimation of the chemical spectra from pollen grains.
Any insights gained from studies on modern grains must also be transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
Fourier transform infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (approximately 1-2 mg) (e.g. @zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), predominantly result with high-quality spectra with detailed biochemical ‘fingerprint’ of pollen, comprising specific signals of sporopollenins, lipids, proteins, carbohydrates and other chemical constituents [@bagciogluMultiscaleVibrationalSpectroscopic2015].
As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017].
However, these approaches may be less relevant for pollen samples extracted from fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.

An alternative approach involves the use of FTIR microscopes, where a FTIR spectrometer is coupled to a microscope. 
This enables acquisition of infrared spectra from a small area, with aperture covering several pollen grains, a single pollen grain, or even pollen grain substructure [@zimmermannVibrationalMicrospectroscopyEnables2015].
FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a; @lukacsRecoveryAbsorbanceSpectra2015].
Although the scattering issues can be tackled by a number of analytical and numerical solutions [@diehnDiscriminationGrassPollen2020; @zimmermannChemicalCharacterizationIdentification2018; @lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016], FTIR measurement of individual pollen grains remains challenging.
Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019]. 
Moreover, microscopic-FTIR techniques have been used to observe chemical differences in pollen samples to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].

One of key challenges faced by chemical palynologists using micro-FTIR and other approaches is the time taken to extract and analyse fossil-pollen samples.
Some FTIR microscopes are equipped with automated sample stages, enabling automated measurement of predetermined measurement positions.
Measurements are usually acquired by consecutive recording of spectra by a single element detector [@zimmermannAnalysisAllergenicPollen2016;@zimmermannChemicalCharacterizationIdentification2018; @zimmermannVibrationalMicrospectroscopyEnables2015].
When spectra are recorded with equidistant spatial steps (i.e. scanning measurement), the individual spectra can be assembled after the recording to obtain a chemical image [@jardineSporopolleninChemistryIts2021; @zimmermannVibrationalMicrospectroscopyEnables2015; @julierChemotaxonomyToolInterpreting2016].
However, recent developments in technologies related to FTIR allow for easier and faster capture of spectra of pollen grains. 
A focal plane array detector (FPA) is made of a 2D array of IR detectors, and allows true chemical imaging.
This enables simultaneous measurement of several single pollen grains from a relatively large surface area without the need to move the sample stage.
Therefore, it is of potentially high relevance for palaeoecological research.

@@

Palaeoecologists may also use a variety of chemical-processing techniques in order to speed-up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of treatments with hydrofluoric acid (HF) and hydrochloric acid (HCl), enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in *Lycopodium* spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample-preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed to investigate the potential effects of the chemical preparation procedures used to isolate pollen grains from different pollen types.

In this study we aimed to explore the use of a FTIR imaging to obtain IR spectra from fossil-pollen *Pinus* grains.
We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the IR spectra from both modern- and fossil-pollen samples.
Pollen was extracted from the sediment cores using two different approaches, namely density separation using sodium polytungstate and acetolysis, followed by FTIR microscopy imaging.
Thus, in addition to testing the feasibility of the FTIR imaging approach on different sediment types, we were were also able to investigate how the IR spectra of the *Pinus* pollen vary as a result of two different extraction procedures.

@@

knitr::kable(sites, booktabs = T, format = "latex", 
             caption = "",
             col.names = c("Name", "Abbr.", "Age", "Description", "Publ.")) %>%
  column_spec(4, width = "5cm")
```

@@

The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
The three sediment cores were selected because they contain *Pinus sylvestris*-pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
*Pinus* pollen in these cores has been described as most likely representing *Pinus sylvestris* in Dalmutladdo and Tiefer See, while it is described as *Pinus* in Meerfelder Maar record.
In this study we focus on *Pinus* pollen for two reasons.
First, *Pinus* spp. is a shade-intolerant, high pollen producer, and so a number of studies have proposed that *Pinus* spp. pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@jardineProxyReconstructionUltravioletB2020; @willisQuantificationUVFlux2011; @seddonPollenchemistryVariationsElevation2021].
Second, although prominent in Quaternary sediment records from Europe, and despite the highest number of species amoing conifers [110 species, @gernandtPhylogenyClassificationPinus2005], *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
Thus, identifying chemical differences in fossil *Pinus* pollen may also allow for identification of crytpic *Pinus* species from the pollen types observed in the fossil record.

Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
First, the samples were extracted using density separation using dissolved sodium polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as a separation medium.
During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
After centrifugation, particulates heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973]. 
In our case the lighter fraction (<1.4 g cm^-3^) contained majority of the pollen grains, after checking both fractions.
Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
Since SPT is a chemically inert salt, it was expected that, after washing, this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.

@@

The spectra were preprocessed before analysis consisted of smoothing and calculation of second derivative and normalization.
We used the Savitzky Golay (SG) algorithm [@savitzkySmoothingDifferentiationData1964] as implemented by the extended multiplicative signal correction (EMSC) package in R [@lilandPackageEMSC2017] and used second degree polynomial and window size five as parameters.
The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
Preprocessing with SG and EMSC has been shown to improve the performance of multivariate analyses, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@woutersenNovelApproachStudy2018; @zimmermannOptimizingSavitzkyGolayParameters2013].

@@

```{r ordinations, echo=FALSE, fig.cap="a. plsr ordination scores all samples diff fresh vs fossil b. regression coef; component 1 c. regression coef component 2; d plsr ordination scores only samples from cores diff spt vs acet; e. regression coef component 1; f. regression coef component 2", message=FALSE, warning=FALSE, out.width="100%"}
knitr::include_graphics(here("figures", "ord_plots.png"))
```

@@

```{r SM, echo=FALSE, fig.cap="Image captured from FTIR microscope of sample slide Dalmutladdo 60 cm depth. Pinus pollen is visible as well as debris and dried residue, presumably BSi.", message=FALSE, warning=FALSE, out.width="90%"}
knitr::include_graphics(here("figures", "SM_BSI.png"))
```

@@

## Extraction of *Pinus* FTIR spectra using the FPA detector
In this study we investigated the use of an FPA detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g. 100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 ?µm? @zimmermannChemicalCharacterizationIdentification2018).
An image with a spectra for each pixel is created, from which spectra of single grains were extracted.

In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1250 - 1000 cm^-1^, and these results build on previous studies which suggest that overall potential for the use of the FPA detector in pollen research **[add references]**.
Thus, moving forward we see potential in the application of the FPA detector in chemical palaeoecology.

@@

Taken together, the differences in the effects of acetolysis across different taxa and sample types suggest that the chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but these results indicate that further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
For example, it would be useful to explore the effects of these extraction methods on modern and fossil material side-by-side, to investigate if there are methods that can bridge the transition from modern to fossil pollen.
Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
This would enable us to compare fossil pollen spectra with fossilized pollen spectra of known species for identification.

@@

The chemical effects of acetolysis that identified across studies suggests that it may be preferable to use an alternative approach to extract fossil pollen grains from sediment samples.
Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], although given the fact that an SPT peak was observed in one of the cores, the results in this paper clearly demonstrate the need to identify efficient dilution processes that may need to be tailored to specific sediment types.
Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973], it is possible that specific extraction densities and procedures may be required from core to core.

@@

## I haven't found a way to integrate these ideas in there. Perhaps we don't need them? I think perhaps we could leave it out
In addition, there is some differentiation between the cores.
Samples from Tiefer See are more distinct from samples of the other two cores, which show more similar scores in the ordination.
One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and one of the main differences between the TSK pollen and the other two locations is sample age.
Whilst the TSK samples are from the past 80 years, MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample likely more than 100 years older than TSK pollen.

This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
Interestingly, @jardineSporopolleninChemistryIts2021 also found differences in FTIR spectra with different sample ages, although they found that that modern-pollen spectra had more in common with spectra from Quaternary pollen, compared to pollen from deeper time samples.
Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.

# References

>>>>

---
title: 'Assessing variations in the chemistry of sub-fossil and modern *Pinus* pollen'
#date: '2020/02/04'
output:
  bookdown::pdf_document2:
    toc: false
    extra_dependencies: ["flafter"]
  bookdown::word_document2:
bibliography: references.bib
csl: biology-letters.csl
---

@@

---
author: `r Authors`
---
`r Affiliations`
<!-- 
2500 words not including in tables title page references
2 figs and 2 tables
figures separate legends in text
separate cover letter

A  200
I  500
MM 400
R  500
D  800
-->

@@

The study of plant microfossils (pollen and spores) is an integral part of palaeoecology and palaeobotany.
Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales [e.g. the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001].

@@

Successful application of pollen-chemistry techniques within palaeoecological research requires a thorough understanding of the factors influencing chemical variations in the modern environment and in the accurate estimation of the chemical spectra from pollen grains.
Any insights gained from studies on modern grains must also be transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
Fourier transform infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (approximately 1-2 mg) (e.g. @zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), predominantly result with high-quality spectra with detailed biochemical ‘fingerprint’ of pollen, comprising specific signals of sporopollenins, lipids, proteins, carbohydrates and other chemical constituents [@bagciogluMultiscaleVibrationalSpectroscopic2015].
As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@zimmermannInfraredSpectroscopyPollen2014; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017; @depciuchIdentificationBirchPollen2018; @muthreichChemicalVariationsQuercus2020; @diehnDiscriminationGrassPollen2020].
However, these approaches may be less relevant for pollen samples extracted from fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.

An alternative approach involves the use of FTIR microscopes, where a FTIR spectrometer is coupled to a microscope. 
This enables acquisition of infrared spectra from a small area, with aperture covering several pollen grains, a single pollen grain, or even pollen grain substructure [@zimmermannVibrationalMicrospectroscopyEnables2015].
FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a].
Although the scattering issues can be tackled by a number of analytical and numerical solutions [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016; @zimmermannChemicalCharacterizationIdentification2018; @diehnDiscriminationGrassPollen2020], FTIR measurement of individual pollen grains remains challenging.
Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018; @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019]. 
Moreover, microscopic-FTIR techniques have been used to observe chemical differences in pollen samples to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016;@jardineSheddingLightSporopollenin2017; @jardineProxyReconstructionUltravioletB2019].

One of key challenges faced by chemical palynologists using micro-FTIR and other approaches is the time taken to extract and analyse fossil-pollen samples.
Some FTIR microscopes are equipped with automated sample stages, enabling automated measurement of predetermined measurement positions.
Measurements are usually acquired by consecutive recording of spectra by a single element detector [@zimmermannVibrationalMicrospectroscopyEnables2015; @zimmermannAnalysisAllergenicPollen2016; @zimmermannChemicalCharacterizationIdentification2018].
When spectra are recorded with equidistant spatial steps (i.e. scanning measurement), the individual spectra can be assembled after the recording to obtain a chemical image [@zimmermannVibrationalMicrospectroscopyEnables2015; @julierChemotaxonomyToolInterpreting2016; @jardineSporopolleninChemistryIts2021].
However, recent developments in technologies related to FTIR allow for easier and faster capture of spectra of pollen grains. 
A focal plane array detector (FPA) is made of a 2D array of IR detectors, and allows true chemical imaging.
This enables simultaneous measurement of several single pollen grains from a relatively large surface area without the need to move the sample stage.
Therefore, it is of potentially high relevance for palaeoecological research.

@@

Palaeoecologists may also use a variety of chemical-processing techniques in order to speed-up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of treatments with hydrofluoric acid (HF) and hydrochloric acid (HCl), enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@schulzeosthoffPhenolsIntegratedCompounds1987; @dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;  @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in *Lycopodium* spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample-preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed to investigate the potential effects of the chemical preparation procedures used to isolate pollen grains from different pollen types.

In this study we aimed to explore the use of a FTIR imaging to obtain IR spectra from fossil-pollen *Pinus* grains.
We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the IR spectra from both modern and fossil pollen samples.
Pollen was extracted from the sediment cores using two different approaches, namely density separation using sodium polytungstate and acetolysis, followed by FTIR microscopy imaging.
Thus, in addition to testing the feasibility of the FTIR imaging approach on different sediment types, we were were also able to investigate how the IR spectra of the *Pinus* pollen vary as a result of two different extraction procedures.

@@

knitr::kable(sites, booktabs = T, format = "latex", 
             caption = "Sampled sediment cores for this study. ",
             col.names = c("Name", "Abbr.", "Age", "Description", "Publ.")) %>%
  column_spec(4, width = "5cm")
```

@@

The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
The three sediment cores were selected because they contain *Pinus sylvestris*-pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
*Pinus* pollen in these cores has been described as most likely representing *Pinus sylvestris* in Dalmutladdo and Tiefer See, while it is described as *Pinus* in Meerfelder Maar record.
In this study we focus on *Pinus* pollen for two reasons.
First, *Pinus* spp. is a shade-intolerant, high pollen producer, and so a number of studies have proposed that *Pinus* spp. pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@willisQuantificationUVFlux2011; @jardineProxyReconstructionUltravioletB2020; @seddonPollenchemistryVariationsElevation2021].
Second, although prominent in Quaternary sediment records from Europe, and despite the highest number of species amoing conifers [110 species, @gernandtPhylogenyClassificationPinus2005], *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
Thus, identifying chemical differences in fossil *Pinus* pollen may also allow for identification of crytpic *Pinus* species from the pollen types observed in the fossil record.

Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
First, the samples were extracted using density separation using dissolved sodium polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as a separation medium.
During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
After centrifugation, particulates heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@flenley1971measurements; @faegriTextbookPollenAnalysis1989; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^ [@juvigneDensiteExinesQuelques1973]. 
In our case the lighter fraction (<1.4 g cm^-3^) contained majority of the pollen grains, after checking both fractions.
Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
Since SPT is a chemically inert salt, it was expected that, after washing, this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.

@@

The spectra were preprocessed before analysis consisted of smoothing and calculation of second derivative and normalization.
We used the Savitzky Golay (SG) algorithm [@savitzkySmoothingDifferentiationData1964] as implemented by the extended multiplicative signal correction (EMSC) package in R [@lilandPackageEMSC2017] and used second degree polynomial and window size five as parameters.
The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
Preprocessing with SG and EMSC has been shown to improve the performance of multivariate analyses, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@zimmermannOptimizingSavitzkyGolayParameters2013; @woutersenNovelApproachStudy2018].

@@

```{r ordinations, echo=FALSE, fig.cap="a. PLSR ordination scores of all samples showing the difference between fresh vs fossil pollen b. regression coefficient of component 1 c. regression coefficient of component 2; d PLSR ordination scores using only samples from sediment cores. Difference between non-acetolysed and acetolyses; e. regression coefficient component 1; f. regression coefficient component 2", message=FALSE, warning=FALSE, out.width="100%"}
knitr::include_graphics(here("figures", "ord_plots.png"))
```

@@

```{r SM, echo=FALSE, fig.cap="Image captured from FTIR microscope of sample slide Dalmutladdo 60 cm depth. \\textit{Pinus} pollen is visible as well as debris and dried residue, presumably BSi.", message=FALSE, warning=FALSE, out.width="90%"}
knitr::include_graphics(here("figures", "SM_BSI.png"))
```

@@

## Extraction of *Pinus* FTIR spectra using the FPA detector
In this study we investigated the use of an FPA detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (500 x 500 µm ), compared to other detectors [e.g. 100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 µm @zimmermannChemicalCharacterizationIdentification2018].
An image with a spectra for each pixel is created, from which spectra of single grains were extracted.

In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@zimmermannCharacterizationPollenVibrational2010; @bagciogluMultiscaleVibrationalSpectroscopic2015] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1250 - 1000 cm^-1^, and these results build on previous studies which suggest that overall potential for the use of the FPA detector in pollen research **[add references]**.
Thus, moving forward we see potential in the application of the FPA detector in chemical palaeoecology.

@@

Taken together, the differences in the effects of acetolysis across different taxa and sample types suggest that chemical extraction methods have an effect on the chemistry of pollen, both modern and fossil origin.
Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but these results indicate that further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
For example, it would be useful to explore the effects of these extraction methods on modern and fossil material side-by-side, to investigate if there are methods that can bridge the transition from modern to fossil pollen.
Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
This would enable us to compare fossil pollen spectra with fossilized pollen spectra of known species for identification.

@@

The chemical effects of acetolysis that identified across studies suggests that it may be preferable to use an alternative approach to extract fossil pollen grains from sediment samples.
Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@regnellPreparativeCentrifugationNew1996; @caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019], although given the fact that an SPT peak was observed in one of the cores, the results in this paper clearly demonstrate the need to identify efficient dilution processes that may need to be tailored to specific sediment types.
Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@flenley1971measurements; @juvigneDensiteExinesQuelques1973; @faegriTextbookPollenAnalysis1989; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998], it is possible that specific extraction densities and procedures may be required from core to core.

@@

<!--
## I haven't found a way to integrate these ideas in there. Perhaps we don't need them? I think perhaps we could leave it out
In addition, there is some differentiation between the cores.
Samples from Tiefer See are more distinct from samples of the other two cores, which show more similar scores in the ordination.
One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and one of the main differences between the TSK pollen and the other two locations is sample age.
Whilst the TSK samples are from the past 80 years, MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample likely more than 100 years older than TSK pollen.

This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
Interestingly, @jardineSporopolleninChemistryIts2021 also found differences in FTIR spectra with different sample ages, although they found that that modern-pollen spectra had more in common with spectra from Quaternary pollen, compared to pollen from deeper time samples.
Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
-->
# References
